A Process for Emulating Comparative Oncology Trials with Real-world Evidence Studies (ENCORE): Process Development and Methodological Considerations for Oncology Real-World Data

### 2025-03-30 manuscript/manuscript.tex
Manuscript updated

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={A Process for Emulating Comparative Oncology Trials with Real-world Evidence Studies},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{A Process for Emulating Comparative Oncology Trials with
Real-world Evidence Studies}
\author{}
\date{}

@@

\textbf{Authors}: Janick Weberpals\textsuperscript{1}, Kenneth L.
Kehl\textsuperscript{2}, Donna R. Rivera\textsuperscript{3}, Pallavi
Mishra-Kalyani\textsuperscript{3}, Georg Hahn\textsuperscript{1},
Priyanka Anand\textsuperscript{1}, Yanina Natanzon\textsuperscript{4},
Andrew J. Belli\textsuperscript{5}, Ching-Kun Wang\textsuperscript{5},
Jenna Collins\textsuperscript{6}\textbf{,}
JonathanKish\textsuperscript{6}\textbf{,} Janet
Espirito\textsuperscript{7}, Nicholas J Robert\textsuperscript{7},
Robert J. Glynn\textsuperscript{1}, Sebastian
Schneeweiss\textsuperscript{1}, Shirley V. Wang\textsuperscript{1}

@@

\textsuperscript{4} ConcertAI, Cambridge, MA, USA

\textsuperscript{5} COTA, Inc., New York, NY, USA

\textsuperscript{6} Flatiron Health, Inc., New York, NY, USA

\textsuperscript{7} Ontada, Boston, MA, USA

@@

\emph{Manuscript last updated: 2025-03-04 18:43:42.372054}

@@

Randomized controlled trials (RCTs) have been the go-to methodology for
establishing the efficacy and safety of medical products. Under the
21\textsuperscript{st} Century Cures Act directive,\textsuperscript{1}
the Food and Drug Administration (FDA) established a framework to
increasingly consider real-world evidence (RWE) generated from
routine-care health data such as electronic health records (EHR) to
evaluate and contextualize the comparative safety and effectiveness of
novel cancer therapies.\textsuperscript{2} Accounting for 21\% of all
approvals, oncology was the disease area with the most FDA drug
approvals in 2023.\textsuperscript{3} RWE has particularly important
potential to complement evidence coming from RCTs in the field of
precision oncology, where potential use cases include the assessment of
effectiveness in specific patient populations that are underrepresented
in RCTs, the construction of external control arms in single-arm trials
where active recruitment to a randomized trial may not be feasible, or
the discovery of biomarkers among pan-tumor populations that harbor
specific genomic and immuno-pathological signatures.

However, the validity and transportability of results derived between
RWE studies and RCTs depends on multiple factors. Frequently referenced
limitations include, lack of baseline randomization, missing data, small
study sizes, data discontinuity,\textsuperscript{4, 5} adoption of
changes in guidelines in real-world care and the inability to measure
and emulate common eligibility criteria, including prognostic factors,
and standardized response assessments in real-world data
(RWD).\textsuperscript{6} While there are already published examples of
oncology trial emulations,\textsuperscript{6--8} a systematic and scaled
approach to emulate a diverse set of different oncology trials with
multiple heterogeneous databases is necessary to gain confidence in the
validity of RWE studies and to provide context as to which questions can
be validly answered.

The RCT DUPLICATE initiative\textsuperscript{9} increased our
understanding of when RWE studies can come to causal conclusions on
treatment effects by benchmarking results against RCTs under the
assumption that each RCT finding reflects a causal treatment effect. In
settings where the RCT designs could be emulated well, RWE studies came
to the same conclusions.\textsuperscript{10} However, prior work from
RCT-DUPLICATE has focused primarily on emulating trials in non-oncology
settings using claims databases.

The \emph{Emulation of Comparative Oncology Trials with Real-world
Evidence} (ENCORE) project\textsuperscript{11} aims to extend this work
to oncology. Studies in oncology come with their own unique set of
challenges which must be systematically explored and understood.
Building on a process co-developed with the FDA through RCT
DUPLICATE,\textsuperscript{9} this expansion to oncology will emulate 12
randomized clinical trials using multiple specialty oncology EHR data
sources. The process will emphasize transparency and include documented
data fitness assessment\textsuperscript{12, 13} for each RWD source with
respect to each trial emulation as well as extensive sensitivity
analyses to assess robustness of findings.

The objectives of this project are to 1) develop state-of-the-art
methodological approaches and 2) to apply them to create insights that
may provide guidance on the potential use of RWE for regulatory science
in oncology. To achieve these objectives, this demonstration project
will systematically emulate 12 oncology trials across four cancers and
assess the agreement of treatment effect estimates between RCTs and
their respective emulations.

@@

The focus of ENCORE is to maximize potential learnings on when RWE
studies can or cannot yield similar results compared to RCTs. To that
end, the emphasis of the project is on trials of therapies for the most
common cancers for which there has been substantial therapeutic
development in recent years. After review and discussion with clinical
and regulatory experts, four cancer indications were identified:
non-small cell lung cancer, breast cancer, colorectal cancer and
multiple myeloma. For each cancer type, we aim to conduct three trial
emulations using multiple databases accessible for the scope of this
project (i.e., the total number of emulations will equal 12 trials x
\emph{n} databases which are found fit-for-purpose for each trial).

We used a semi-automated process for trial selection where the
eligibility criteria are documented in CONSORT diagrams showing reasons
for excluding RCTs for each cancer type. The search was conducted using
the \emph{Aggregated Analysis of ClinicalTrials.gov} (AACT) database
which is a publicly available relational database developed and
maintained by the Clinical Trials Transformation Initiative (CTTI) that
contains all information (protocol and result data elements) about every
study registered on ClinicalTrials.gov.\textsuperscript{14} To identify
eligible trials, we used a combined search query strategy of the
National Library of Medicine (NLM)-controlled \emph{Medical Subject
Headings \textbf{(}}MeSH) term and a free keyword search for the
respective cancer indication in the \emph{conditions}, \emph{studies}
and \emph{detailed\_descriptions} fields of each trial entry on
ClinicalTrials.gov.

@@

\begin{itemize}
\item
  Interventional
\item
  Randomized
\item
  Intervention model: parallel assignment
\item
  Industry-sponsored
\item
  Trial start in 2011 or later
\item
  Primary purpose was to study treatment effects
\item
  Overall survival must be one of the endpoints reported (either as
  hazard ratio or median overall survival time)
\item
  Recruitment status: `Completed' or `Active, not recruiting'
\item
  Feasibility and clinical relevance (latter was defined as treatment or
  paradigm-changing trials or such that challenged existing treatment
  policies)
\end{itemize}

The rationale and operationalization of each criterion is listed in
detail in Table~\ref{tbl-criteria}. We will mainly consider pivotal
interventional, randomized trials after 2011 because treatment
guidelines among included cancer indications have undergone significant
changes in recent years. Due to the rapid adoption of new breakthrough
therapies in routine care, it is unlikely to find patients who may be
still treated with outdated treatment regimens in current clinical
practice. Conversely, we excluded trials with results published too
recently. To allow for enough data and follow-up time accrual in the
databases used for this project. Although there have been substantial
methodological advancements to increase our understanding on the
emulation and comparison of real-world progression-free survival (PFS)
and objective response rates (ORR) to a
RECISTv1.1\textsuperscript{15}-based PFS and ORR assessment in
RCTs\textsuperscript{16, 17}, imaging-based evaluations still hold a
level of granularity which may not be necessarily reflected in
chart-abstracted assessments of a patient's progression in routine
care.\textsuperscript{18, 19} The timing and cadence of intervals
between progression assessments can differ between RCTs and routine care
which may result in measurement error and bias.\textsuperscript{20}
Given this, and the large number of other methodological challenges like
missing data, small sample sizes, data discontinuity and rapidly
changing guideline treatments, we focus on the emulation of overall
survival (OS) as the endpoint of interest. Therefore, we include only
trials that have reported overall survival (OS) as one of the
pre-specified endpoints in the protocol.

While most trial eligibility criteria could be operationalized in an
automated fashion, our final criterion was an assessment of emulation
feasibility and clinical relevance. This criterion involved extensive
human review. The critical points considered in this step include an
initial feasibility assessment of the data fitness,\textsuperscript{19}
including assessment of whether critical eligibility criteria (e.g.,
biomarker status) and prognostic factors (e.g., ECOG performance score)
are measurable and whether preliminary study size counts are reasonable.
Lastly, trial candidates were ranked and shortlisted into primary and
runner-up candidates based on their clinical and regulatory relevance.

A list of tentative, shortlisted primary candidates is presented in
Table~\ref{tbl-rcts} and the corresponding selection process is
illustrated in the CONSORT diagrams (Supplementary Figures 1-4).
Naturally, the majority of trials cover advanced (locally advanced,
inoperable, recurrent/progressive disease) or metastatic cancer
populations because a large proportion of drug development efforts start
in these settings. A key objective that we aim to foster with the
shortlisted trials is to achieve a better understanding how different
disease settings (early, late), line settings ({[}neo{]}adjuvant, first
line, advanced lines of therapy), therapy protocols (monotherapy,
combination therapy) and population characteristics (simple versus
complex genetic or immunological signatures) can be emulated using RWD.
If more thorough feasibility assessments suggest that the threat of bias
from mis-measurement of key study parameters or residual confounding
remains high, or that the study size is not sufficient, then runner-up
candidates will be considered instead.

@@

The ENCORE project will use data from four oncology-specific electronic
health records (EHR)-derived data sources (in alphabetical order):
ConcertAI, COTA, Flatiron Health, and Ontada/McKesson. All available
databases draw from a comprehensive national sample of patients with
cancer in the US with the detailed EHR-based data necessary to study
medication effectiveness in oncology. A detailed description and
methodology on how patients are sampled will be provided with each trial
emulation protocol. For ENCORE, not all databases will be available for
each cancer indication and the names of the databases will be blinded
and referred to as ENCORE DataBase (EDB) 1, 2, 3 and 4 for the final
reporting of results (the numbering does not coincide with the above
order of mention of the databases). If more than one database is
considered fit-for-purpose for a respective trial emulation, the best
possible analytic model will be employed for each database separately
and final treatment effect estimates will be pooled using a
meta-analytic approach using fixed effects (weights reflect the inverse
of the variance of each database study) and random effects (weights
reflect the inverse of the variance of each database study plus an
additional variance term that quantifies the assumed heterogeneity
between databases) models.\textsuperscript{21}

@@

For each shortlisted and selected RCT, a detailed protocol,
pre-specifying key elements of the trial emulation, will be developed
using the HARPER protocol template\textsuperscript{22} recommended for
regulatory submissions of RWE studies\textsuperscript{23} and will be
registered on ClinicalTrials.gov after review by a clinical and FDA
regulatory expert panel. Following the target trial emulation framework,
we will provide an explicit statement and rationale on how each element
will be emulated including database selection, covariate measurement,
operationalization of key eligibility criteria, study design, data
analysis and causal contrasts of interest.\textsuperscript{24, 25} Since
it is common that oncology RCTs update OS estimates periodically based
on accrued follow-up time, the protocol will give a brief summary of
each emulated RCT and specify which target OS estimates will be used to
compare agreement metrics to (see Section~\ref{sec-agreement-metrics}).
All eligibility criteria will be extracted based on publications,
publicly available protocols and statistical analysis plans of the
selected RCT.

@@

Real-world data fitness and emulation feasibility for each shortlisted
candidate trial will be assessed in multiple steps based on guidance of
the oncology quality, characterization, and assessment of real-world
data (Oncology QCARD) Initiative.\textsuperscript{19} The first step
assesses if relevant variables like exposure/line of therapy, outcomes,
and covariates are generally available, measured and operationalizable
in routine-care. Since many oncological RCTs in recent years have
focused on selected, biomarker-defined populations, subtleties in
measurement and operationalizability of specific biomarkers must be
reflected to ensure a representative and large enough study population.
For example, immune checkpoint inhibitors have significantly changed the
cancer treatment landscape since the first approvals of the CTLA-4
inhibitor ipilimumab in 2011 and the PD-1 inhibitors pembrolizumab and
nivolumab in 2014 in the United States. With many trials that have
followed thereafter, the operationalization of the expression of the
PD-L1 biomarker in RCTs (e.g., as a percent staining, tumor proportion
score or combined positive score) has also evolved, and PD-L1
'\emph{positivity'} may have different definitions across calendar years
based on different cut-off values.

According to the \emph{Structured Process to Identify Fit-For-Purpose
Data} (SPIFD) framework\textsuperscript{13}, the next step will outline
tables that describe how eligibility criteria will be ascertained using
a color-coded heatmap that will indicate the level of confidence on how
well each criterion can be emulated in each selected database. As there
are general eligibility criteria in oncology trials which either will
not be possible to emulate (e.g., physician-assessed survival prognosis
of x months) or that are clinically not relevant for the emulation of
the trial (e.g., male patients should be willing to use barrier
contraception), the study team will decide on key eligibility criteria
for the emulation of the trial.

We will additionally provide a definition on how exposures, outcomes and
covariates are exactly defined and operationalized in each respective
database. There will be special emphasis on how exposure, in context of
their respective disease and line of therapy settings, and the OS
outcome will be emulated. For all considered databases, the OS endpoint
is typically a composite that, depending on the underlying database, can
be derived from different sources comprising EHR documentation, social
security death index, obituary and other linkages. Given that not all
relevant sources that provide mortality data are synchronized and
updated uniformly, sensitivity analyses with more conservative (i.e.,
earlier) censoring dates will be considered for each trial emulation to
mitigate the potential impact of ghost-time bias.\textsuperscript{26}

@@

Critical aspects when emulating oncology trials are the choice and
estimation of the appropriate estimand of interest.\textsuperscript{27}
Particularly when emulating pivotal trials of paradigm-changing
treatments, multiple aspects need to be considered such as the
contemporaneity of the (historical) control cohort, the adoption rate of
the novel intervention in routine care, the magnitude of the clinical
treatment benefit and the rate in which (particularly patients in the
control arm) discontinue or cross-over to the interventional treatment,
which could lastly bias emulated treatment effects towards the null. To
that end, comprehensive data explorations will be performed as part of
the protocol development to contextualize these parameters and (if
reported) draw comparisons to the emulated trial. Examples for such
standard diagnostics are visualized in Figure~\ref{fig-initiators}. All
exploratory analyses will be conducted blinded towards the outcome to
avoid influencing study design and analytic choices.

The distribution of patient characteristics, stratified by exposure
status, will be examined in Table 1's before and after applying
eligibility criteria and contrasted with the distributions of patient
characteristics of the original RCT. Initial propensity score matching
or weighting methods will be applied to ensure that measured
pre-exposure covariates can be balanced, exposure cohorts are
conditionally exchangeable at baseline and resulting sample sizes are
still sufficient after matching or weighting. At this stage, all
exploratory analyses will be conducted blinded towards the outcome to
not bias any study design and analytic choices based on known outcome
information.

@@

Causal analyses of observational data may not have the same
pre-requisites in terms of formal hypothesis testing and statistical
power as RCTs since the number of `recruited' patients is given and
cannot be influenced.\textsuperscript{28} For this project, however, the
emulations in EHR data must have at least equal power to the relevant
trial for interpretation of the pre-specified agreements metrics between
RCT and RWD results. Since the main outcome of interest is defined as
time to all-cause mortality (OS), the estimation of the statistical
power is driven by the number of events rather than the number of
patients. To assess if the overall number of events, unstratified by
exposure, is sufficient such that a significant difference can be
detected based on the original RCT-reported hazard ratio (HR), the
statistical power (1-\(\beta\)) will be estimated using Schoenfeld's
sample-size formula for the proportional-hazards regression
model.\textsuperscript{29}

@@

To establish an analytic cohort, key eligibility criteria will be
applied in which patients with missing values in eligibility criteria
are considered eligible in the respective attrition steps to allow for
thorough missing data investigations. These missing data investigations
will empirically assess assumptions on potential underlying missingness
mechanisms according to Rubin's classification of missing data (i.e.,
missing completely at random {[}MCAR{]}, missing at random {[}MAR{]} and
missing not at random {[}MNAR{]}).\textsuperscript{32} To that end, we
will adopt a principled process on missing data that empirically
evaluates different aspects across partially observed covariates based
on three group diagnostics.\textsuperscript{33, 34} These diagnostics
cover (1) comparisons of patients characteristics with and without an
observed level of the partially observed covariate, (2) ability to
predict missingness given observed data, and (3) assessments if outcomes
between patients with a missing value are systematically different.
Together with expert domain knowledge and assumptions about the
underlying missing data structure through canonical causal
diagrams,\textsuperscript{35} this will inform decisions regarding the
in- or exclusion of patients with missing values in key eligibility
criteria and potential sensitivity analyses to assess the robustness of
these decisions.\textsuperscript{36}

While the MAR assumption is a strong assumption to hold across all
considered covariates, it was shown that especially in the context of
partially observed covariate data (as opposed to missing exposure and
outcome data), only mechanisms in which a covariate causes its own
missingness leads to critical bias (MNAR).\textsuperscript{35} Hence,
methodologies which retain patients and give the potential to adjust for
a broader set of prognostic factors (e.g., multiple
imputation\textsuperscript{37} or doubly robust
methods\textsuperscript{38}) may be preferred over complete case
analyses.

\subsubsection{Endpoints and propensity score
analyses}\label{endpoints-and-propensity-score-analyses}

Due to its ubiquity in oncology trials, the primary parameter of
interest in ENCORE will be defined as the marginal hazard ratio (HR)
coefficient for the treatment comparison for time to all-cause mortality
(OS).\textsuperscript{39} We will also consider alternative endpoints on
an absolute risk scale such as median survival times, survival
probabilities at pre-defined time points during
follow-up\textsuperscript{40} and restricted mean survival times as
secondary endpoints of interest.

@@

\caption{\label{tbl-metrics}Visualization of agreement metrics example.}

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={A Process for Emulating Comparative Oncology Trials with Real-world Evidence Studies (ENCORE): Process Development and Methodological Considerations for Oncology Real-World Data},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{A Process for Emulating Comparative Oncology Trials with
Real-world Evidence Studies (ENCORE): Process Development and
Methodological Considerations for Oncology Real-World Data}
\author{}
\date{}

@@

\textbf{Authors}: Janick Weberpals\textsuperscript{1}, Kenneth L.
Kehl\textsuperscript{2}, Donna R. Rivera\textsuperscript{3}, Pallavi
Mishra-Kalyani\textsuperscript{3}, Catherine C.
Lerro\textsuperscript{3}, Erin Larkins\textsuperscript{4}, Sonia
Singh\textsuperscript{4}, Preeti Narayan\textsuperscript{4}, Richard
Curley\textsuperscript{4}, Paul Kluetz\textsuperscript{4}, Georg
Hahn\textsuperscript{1}, Priyanka Anand\textsuperscript{1}, Yanina
Natanzon\textsuperscript{5}, Andrew J. Belli\textsuperscript{6},
Ching-Kun Wang\textsuperscript{6}, Jenna
Collins\textsuperscript{7}\textbf{,} Jonathan
Kish\textsuperscript{7}\textbf{,} Janet Espirito\textsuperscript{8},
Nicholas J Robert\textsuperscript{8}, Robert J.
Glynn\textsuperscript{1}, Sebastian Schneeweiss\textsuperscript{1},
Shirley V. Wang\textsuperscript{1}

@@

\textsuperscript{4} Center for Drug Evaluation and Research, US Food and
Drug Administration, Silver Spring, MD, USA

\textsuperscript{5} ConcertAI, Cambridge, MA, USA

\textsuperscript{6} COTA, Inc., New York, NY, USA

\textsuperscript{7} Flatiron Health, Inc., New York, NY, USA

\textsuperscript{8} Ontada, Boston, MA, USA

@@

\emph{Manuscript last updated: 2025-03-30 15:14:45.088413}

@@

Randomized controlled trials (RCTs) are the go-to methodology for
establishing the efficacy and safety of medical products. Under the
21\textsuperscript{st} Century Cures Act,\textsuperscript{1} the Food
and Drug Administration (FDA) established a framework to factor
consideration of real-world evidence (RWE) generated from routine-care
health data such as electronic health records (EHR) to evaluate and
contextualize the safety and effectiveness of novel cancer
therapies.\textsuperscript{2} Accounting for 30\% of new drug approvals,
oncology was the disease area with the most FDA drug approvals in
2024\textsuperscript{3} and has many areas of high unmet medical need to
treat serious conditions. Therefore, RWE has particularly important
potential to complement evidence coming from RCTs in the field of
oncology. Potential use cases include the assessment of effectiveness in
specific patient populations that are underrepresented in RCTs, the
construction of external control arms in single-arm trials that have
feasibility or equipoise challenges, or the precision oncology-focused
discovery of biomarkers among pan-tumor populations that harbor specific
genomic and immuno-pathological signatures.

However, the validity and transportability of results derived between
RWE studies and RCTs depends on multiple factors. Frequently referenced
limitations include, lack of baseline randomization, missing data, small
study sizes, data discontinuity,\textsuperscript{4, 5} adoption of
changes in guidelines in real-world care and the inability to measure
and emulate common eligibility criteria, including prognostic factors,
and standardized response assessments in real-world data
(RWD).\textsuperscript{6} While examples of oncology trial emulations
have been published,\textsuperscript{6--8} a systematic and scaled
approach to emulate a diverse set of different oncology trials with
multiple heterogeneous databases is necessary to gain confidence in the
validity of RWE studies, evaluate regulatory considerations, and to
provide context as to which questions can be validly answered.

The RCT DUPLICATE initiative\textsuperscript{9} increased our
understanding of when RWE studies can come to causal conclusions on
treatment effects by benchmarking results against RCTs under the
assumption that each RCT finding reflects a causal treatment effect. In
settings where the RCT designs could be emulated well, RWE studies were
able to reach the same conclusions.\textsuperscript{10} However, this
prior work from RCT-DUPLICATE focused primarily on emulating trials in
non-oncology settings using claims databases.

The \emph{Emulation of Comparative Oncology Trials with Real-world
Evidence} (ENCORE) project\textsuperscript{11} aims to extend this work
to oncology. Studies in oncology come with their own unique set of
methodological challenges which must be systematically explored and
understood. Building on a process co-developed with the FDA through RCT
DUPLICATE,\textsuperscript{9} this expansion to oncology will emulate 12
randomized clinical trials using multiple specialty oncology EHR data
sources. The process will emphasize transparency and include documented
data fit-for-purpose assessment\textsuperscript{12, 13} for each RWD
source with respect to each trial emulation as well as extensive
sensitivity analyses to assess robustness of findings.

The objectives of this project are to 1) develop state-of-the-art
methodological approaches and 2) to apply them to create insights into
the potential use of RWE to enhance regulatory science in oncology. To
achieve these objectives, this demonstration project will systematically
emulate 12 oncology trials across four cancer types and assess the
agreement of treatment effect estimates between RCTs and their
respective emulations.

@@

The focus of ENCORE is to efficiently explore in what contexts RWE
studies can or cannot yield similar results compared to RCTs. Therefore,
the project emphasizes trials of therapies for common cancers for which
there has been substantial therapeutic development in recent years.
After review and collaboration with clinical and regulatory experts,
four cancer indications were identified: non-small cell lung cancer,
breast cancer, colorectal cancer and multiple myeloma. For each cancer
type, we aim to conduct three trial emulations using multiple accessible
databases (i.e., the total equaling 12 trial emulations x \emph{n}
databases which are found fit-for-purpose for each trial).

We used a systematic process for trial selection where the eligibility
criteria are documented in CONSORT diagrams showing reasons for
excluding RCTs for each cancer type. The search was conducted using the
\emph{Aggregated Analysis of ClinicalTrials.gov} (AACT) database which
is a publicly available relational database developed and maintained by
the Clinical Trials Transformation Initiative (CTTI) that contains
information (protocol and result data elements) about studies registered
on ClinicalTrials.gov.\textsuperscript{14} To identify eligible trials,
we used a combined search query strategy of the National Library of
Medicine (NLM)-controlled \emph{Medical Subject Headings
\textbf{(}}MeSH) term and a free keyword search for the respective
cancer indication in the \emph{conditions}, \emph{studies} and
\emph{detailed\_descriptions} fields of each trial entry on
ClinicalTrials.gov.

@@

\begin{itemize}
\item
  Interventional
\item
  Randomized
\item
  Intervention model: parallel assignment
\item
  Industry-sponsored
\item
  Trial start in 2011 or later
\item
  Primary purpose was study treatment efficacy
\item
  Overall survival must be one of the endpoints reported (either as
  hazard ratio or median overall survival time)
\item
  Recruitment status: `Completed' or `Active, not recruiting'
\item
  Feasibility and clinical relevance (latter was defined as treatment or
  paradigm-changing trials or such that challenged existing treatment
  policies)
\end{itemize}

The rationale and operationalization of each criterion is listed in
detail in Table~\ref{tbl-criteria}. We will mainly consider pivotal
large late-phase interventional RCTs after 2011 because treatment
guidelines among included cancer indications have undergone significant
changes in recent years. Due to the rapid adoption of breakthrough
therapies in routine care, it is unlikely to find patients who may be
still treated with outdated treatment regimens in current clinical
practice. Conversely, we excluded trials with results published too
recently. To allow for enough data and follow-up time accrual in the
real-world databases used for this project. Although there have been
substantial methodological advancements to increase our understanding on
the emulation and comparison of real-world progression-free survival
(PFS) and objective response rates (ORR) to a
RECISTv1.1\textsuperscript{15}-based PFS and ORR assessment in
RCTs\textsuperscript{16, 17}, imaging-based evaluations still hold a
level of granularity which may not is often not reflected in
chart-abstracted assessments of a patient's progression in routine
care.\textsuperscript{18, 19} The timing and cadence of intervals
between progression assessments can differ between RCTs and routine care
which may result in measurement error and bias.\textsuperscript{20}
Given this, and the large number of other methodological challenges
including missing data, small sample sizes, data discontinuity and
rapidly changing guideline treatments, we focus on the emulation of
overall survival (OS) as the endpoint of interest. Therefore, we include
only trials that have reported overall survival (OS) as one of the
pre-specified endpoints in the protocol.

While most trial eligibility criteria could be operationalized in an
automated fashion, our final criterion was an assessment of emulation
feasibility and clinical relevance. This criterion involved extensive
human review. The fundamental points considered in this step include an
initial feasibility assessment of the data fitness,\textsuperscript{19}
including assessment of whether critical eligibility criteria (e.g.,
biomarker status) including prognostic factors (e.g., ECOG performance
score) are measurable and whether preliminary study size is reasonable.
Lastly, trial candidates were ranked and shortlisted into primary and
runner-up candidates based on their clinical and regulatory relevance.

A list of tentative, shortlisted primary candidates is presented in
Table~\ref{tbl-rcts} and the corresponding selection process is
illustrated in the CONSORT diagrams (Supplementary Figures 1-4).
Naturally, the majority of trials cover advanced (locally advanced,
inoperable, recurrent/progressive disease) or metastatic cancer
populations because a large proportion of drug development efforts have
recently focused on these settings. A key objective that we aim to
explore with the selected trials is to achieve a better understanding
how different disease settings (early, late), line settings
({[}neo{]}adjuvant, first line, advanced lines of therapy), therapy
protocols (monotherapy, combination therapy) and clinical
characteristics (all-comer versus complex genetic or immunological
signatures) can be emulated using RWD. If more thorough feasibility
assessments suggest that the threat of bias from mis-measurement of key
study parameters or residual confounding remains high, or that the study
size is not sufficient, then runner-up candidate trials will be
considered instead.

@@

The ENCORE project will use data from four oncology-specific electronic
health records (EHR)-derived RWD data sources (in alphabetical order):
ConcertAI, COTA, Flatiron Health, and Ontada/McKesson. The included
databases comprise the necessary information to study medical product
effectiveness in oncology derived from US-based oncology EHRs. A
detailed description and sampling methodology will be provided with each
trial emulation protocol. For ENCORE, not all databases will be
available for each cancer indication and the names of the databases will
be blinded and referred to as ENCORE DataBase (EDB) 1, 2, 3 and 4 for
the final reporting of results (the numbering does not coincide with the
above order of mention of the databases). If more than one database is
considered fit-for-purpose for a respective trial emulation, the best
possible analytic model will be employed for each database separately
and final treatment effect estimates will be pooled using a
meta-analytic approach using fixed effects (weights reflect the inverse
of the variance of each database study) and random effects (weights
reflect the inverse of the variance of each database study plus an
additional variance term that quantifies the assumed heterogeneity
between databases) models.\textsuperscript{21}

@@

For each selected RCT, a detailed protocol, pre-specifying key elements
of the trial emulation, will be developed using the HARPER protocol
template\textsuperscript{22} recommended for regulatory submissions of
RWE studies\textsuperscript{23} and will be registered on
ClinicalTrials.gov after review by a clinical and FDA regulatory expert
panel. Following the target trial emulation framework, we will provide
an explicit statement and rationale on how each element will be emulated
including database selection, covariate measurement, operationalization
of key eligibility criteria, study design, data analysis and causal
contrasts of interest.\textsuperscript{24, 25} Since it is common that
oncology RCTs update OS estimates periodically based on accrued
follow-up time, the protocol will give a brief summary of each emulated
RCT and specify which target OS estimates will be used for agreement
metric evaluation (see Section~\ref{sec-agreement-metrics}). All
eligibility criteria will be extracted based on publications, publicly
available protocols, and statistical analysis plans of the selected RCT.

@@

Real-world data fitness and emulation feasibility for each shortlisted
candidate trial will be assessed in multiple steps based on direction
from of the oncology quality, characterization, and assessment of
real-world data (Oncology QCARD) Initiative.\textsuperscript{19} The
first step assesses if relevant variables like exposure, line of
therapy, outcomes, and covariates are generally available, measured and
operationalizable in routine-care. Since many oncological RCTs in recent
years have focused on selected, biomarker-defined populations, nuances
in measurement and operationalizability of specific biomarkers must be
reflected to ensure a representative and large enough study population.
For example, immune checkpoint inhibitors have significantly changed the
cancer treatment landscape since the first approvals of the CTLA-4
inhibitor ipilimumab in 2011 and the PD-1 inhibitors pembrolizumab and
nivolumab in 2014 in the United States. However, the operationalization
of the expression of the PD-L1 biomarker in RCTs (e.g., as a percent
staining, tumor proportion score or combined positive score) has
evolved, and as such PD-L1 '\emph{positivity'} may have different
definitions across calendar years based on different cut-off values.

According to the \emph{Structured Process to Identify Fit-For-Purpose
Data} (SPIFD) framework\textsuperscript{13}, the next step outlines how
eligibility criteria will be ascertained using a color-coded heatmap
that will indicate the level of confidence on how well each criterion
can be emulated in each selected database. As there are general
eligibility criteria in oncology clinical trials which either are
infeasible to emulate (e.g., physician-assessed survival prognosis of x
months) or that are clinically not relevant for the emulation of the
trial (e.g., male patients should be willing to use barrier
contraception), the study team will decide on key eligibility criteria
for the emulation of the trial.

Additionally, we will provide conceptual and operationalized definitions
on how exposures, outcomes and covariates are defined in each respective
database. There will be special emphasis on how exposure, in context of
disease and line of therapy settings, and the OS outcome are emulated.
For all considered databases, OS is typically a composite measure that,
depending on the underlying database, can be derived from different
mortality sources (i.e., EHR documentation, social security death index,
obituary and other linkages). Given that not all relevant sources that
provide mortality data are synchronized and updated uniformly,
sensitivity analyses with more conservative (i.e., earlier) censoring
dates will be considered for each trial emulation to mitigate the
potential impact of ghost-time bias.\textsuperscript{26}

@@

One important aspect when emulating oncology trials is the choice and
estimation of the appropriate estimand of interest.\textsuperscript{27}
Particularly when emulating pivotal trials of paradigm-changing
treatments, multiple aspects need to be considered such as the
contemporaneity of the control cohort, the adoption rate of the novel
medical product in routine care, the magnitude of the clinical treatment
benefit and the rate in which (particularly patients in the control arm)
discontinue or cross-over to the interventional treatment, which could
lastly bias emulated treatment effects towards the null. To that end,
comprehensive data explorations will be performed as part of the
protocol development to contextualize these parameters and (if reported)
draw comparisons to the emulated trial. Examples for such standard
diagnostics are visualized in Figure~\ref{fig-initiators}. All
exploratory analyses will be conducted blinded towards the outcome to
avoid influencing study design and analytic choices.

The distribution of patient characteristics, by exposure status, will be
examined using Table 1's before and after applying eligibility criteria
and contrasted with the distributions of patient characteristics of the
original RCT. Initial propensity score matching or weighting methods
will be applied to ensure that measured pre-exposure covariates can be
balanced, exposure cohorts are conditionally exchangeable at baseline
and resulting sample sizes are still sufficient after matching or
weighting. At this stage, all exploratory analyses will be conducted
blinded towards the outcome to not bias any study design or analytic
choices based on known outcome information.

@@

Causal analyses of observational data are often not designed with
respect to formal hypothesis testing and statistical power in the same
manner as RCTs since the number of `recruited' patients is limited by
available data and cannot be influenced.\textsuperscript{28} For this
project, however, the emulations in EHR data must have at least equal
power to the relevant trial for interpretation of the pre-specified
agreements metrics between RCT and RWD results. Since the main outcome
of interest is defined as time to all-cause mortality (OS), the
estimation of the statistical power is driven by the number of events
rather than the number of patients. To assess if the overall number of
events, unstratified by exposure, is sufficient such that a significant
difference can be detected based on the original RCT-reported hazard
ratio (HR), the statistical power (1-\(\beta\)) will be estimated using
Schoenfeld's sample-size formula for the proportional-hazards regression
model.\textsuperscript{29}

@@

To establish an analytic cohort, missingness will be assessed in a first
iteration across patients who meet all eligibility criteria, including
those with missing values. These missing data investigations will
empirically assess assumptions on potential underlying missingness
mechanisms according to Rubin's classification of missing data (i.e.,
missing completely at random {[}MCAR{]}, missing at random {[}MAR{]} and
missing not at random {[}MNAR{]}).\textsuperscript{32} We will adopt a
principled process on missing data that empirically evaluates different
aspects across partially observed covariates based on three group
diagnostics.\textsuperscript{33, 34} These diagnostics cover (1)
comparisons of patients characteristics with and without an observed
level of the partially observed covariate, (2) ability to predict
missingness given observed data, and (3) assessments if outcomes between
patients with a missing value are systematically different. Expert
domain knowledge and assumptions about the underlying missing data
structure through canonical causal diagrams\textsuperscript{35} will
additionally inform decisions regarding the inclusion or exclusion of
patients with missing values in key eligibility criteria and potential
sensitivity analyses to assess the robustness of these
decisions.\textsuperscript{36}

While the MAR assumption is a strong assumption to hold across all
considered covariates, it has been shown that especially in the context
of partially observed covariate data (as opposed to missing exposure and
outcome data), only mechanisms in which a covariate causes its own
missingness lead to critical bias (MNAR).\textsuperscript{35} Hence,
methodologies which retain patients and give the potential to adjust for
a broader set of prognostic factors (e.g., multiple
imputation\textsuperscript{37} or doubly robust
methods\textsuperscript{38}) may be preferred over complete case
analyses.

\subsubsection{Outcome and propensity score
analyses}\label{outcome-and-propensity-score-analyses}

Due to its ubiquity in oncology trials, the primary parameter of
interest in ENCORE will be defined as the marginal hazard ratio (HR)
coefficient for the treatment comparison for OS.\textsuperscript{39} We
will also consider alternative endpoints on an absolute risk scale such
as median survival times, survival probabilities at pre-defined time
points during follow-up\textsuperscript{40} and restricted mean survival
times as secondary endpoints of interest.

@@

\caption{\label{tbl-metrics}Example visualization of agreement metrics.}
